Gesture Typing on Virtual Tabletop: Effect of Input on Performance

### 2017-08-22 paper.tex
up to discussion, missing conclusion, trim citations format

\numberofauthors{4}
% Notice how author names are alternately typesetted to appear ordered in 2-column format;
% i.e., the first 4 autors on the first column and the other 4 auhors on the second column.
% Actually, it's up to you to strictly adhere to this author notation.
\author{
  \alignauthor{
    \textbf{Antoine Loriette}\\
    \affaddr{AuthorCo, Inc.}\\
    \affaddr{123 Author Ave.}\\
    \affaddr{Authortown, PA 54321 USA}\\
    \email{antoine.loriette@gmail.com}
  }\alignauthor{
    \textbf{Sebastian Stein}\\
    \affaddr{AuthorCo, Inc.}\\
    \affaddr{123 Author Ave.}\\
    \affaddr{Authortown, PA 54321 USA}\\
    \email{author5@anotherco.com}
  }
  \vfil
  \alignauthor{
    \textbf{Roderick Murray-Smith}\\
    \affaddr{AuthorCo, Inc.}\\
    \affaddr{123 Author Ave.}\\
    \affaddr{Authortown, PA 54321 USA}\\
    \email{author2@anotherco.com}
  }\alignauthor{
    \textbf{John H. Williamson}\\
    \affaddr{AuthorCo, Inc.}\\
    \affaddr{123 Author Ave.}\\
    \affaddr{Authortown, PA 54321 USA}\\
    \email{author6@anotherco.com}
  }
\vfil
}

@@

\begin{abstract}
The association of tabletop interaction with gesture typing presents potential for physically impaired users. In this work, we use depth cameras to create touch surfaces on regular tabletops. We describe our prototype system and report on a supervised learning approach to fingertips touch classification. We follow with a gesture typing study that compares our system with a control tablet scenario and explore the influence of input size and aspect ratio of the virtual surface on the text input performance. We show that novice users perform with the same error rate at half the input rate with our system as compared to the control condition, that an input size between A5 and A4 present the best tradeoff between precision, performance and user preference and that user's indirect tracking ability seems to be the overall performance limiting factor.

@@

Depth cameras~\cite{Harrison2011,Xiao2016} have usually been employed to create such surfaces, even though touch classification remains problematic and few systems are available for all camera types. Other work have tried to combine different sensor sources (cite Ubitouch) or other techniques (FTIR?).

@@

% =============================================================================
\section{System overview}
% =============================================================================
The envisioned interaction scenario is shown in ~\autoref{fig:banner} with its implementation in \autoref{fig:system_insitu_0}. We use an Intel Realsense SR300 obliquely mounted on a tripod that overlooks the interaction surface. Finger detection and tracking is performed on a desktop computer. The system implements the 3 state button model, with its touch events sent to the Android tablet running a custom application overwriding its input event system. In addition, hover state is displayed as a red marker at the pointer’s position, touching is displayed through the Android debugging facility as a cross spanning the screen and continuous touch is displayed as a trace. An audio feedback is produced on each touch down event.

@@

The touch classification problem is usually solved via flood-filling~\cite{Harrison2011,Xiao2016}. However, it is well suited to a supervised learning approach provided availability of a training dataset. One of the authors video-taped 6 minutes of interactions, half touching and half hovering with some frames out of range while using the index, thumb and pinky at close and long range. From the fingertip pointcloud, we use the histogram of z-values as features (providing some orientation and shape invariance) and a neural network classifier. We tested the generalisation with different cross validation splits and obtained an averaged 0.96 AUC for the ROC, see~\autoref{fig:roc_auc}. The ROC curves led us to choose an operating point of 0.5 for the live system, which we trained on the whole dataset for 75 epochs.

@@

% =============================================================================
\section{User study}
% =============================================================================
We ran an experiment with two research goals in mind. First we wanted to evaluate the usability of our system, for this purpose we included a condition with an interaction directly on the tablet. Second, given the nature of the task and its inherent difficulty for novice users, we investigated the influence of the physical space on performance (Accot et al.~\cite{Accot2001} showed a U-shaped curve in performance). Our independent variable was thus the control surface dimensions, which is effectively changing the control-display gain defined by: $CD_{gain} = V_{display}/V_{control}$.

@@

We gathered 12 participants, all right handed, without requirement on gesture typing experience. A repeated measures within-subjects design was used. There were 3 conditions and 5 level combinations were evaluated, see \autoref{tab:cdt} for details. The presentation order of the 5 combinations was randomised and presented a uniform distribution across the participants. The task was to write 20 words per level (words taken at random from the most common english words with length between 2 and 5 letters) with maximum 7 attempt to complete 5 correct input for each word. The design of the task is similar to \cite{Quinn2016} and allows novice users to focus on the physical execution instead of the shape recollection, effectively emulating a proficient bahavior even for novice users. After each level, participants were offered to take a break before moving to the next one. Finally, participants were asked for their feedback using the NASA Task Load Index~\cite{Hart1988} (NASA-TLX) to assess the perceived workload of completed level.

@@

% =============================================================================
\section{Result}
% =============================================================================
The dependent variables are the success rate, the time taken per trial and the trace data when available. This allowed us to compute the dependent measure error rate defined as the percentage of unsucessful attempts as well as the text entry rate measured in Words Per Minute (WPM), as in [3], according to the formula\footnote{$WPM = |T|/s \times 60/5$ where $|T|$ is the length of the transcribed string, s is time in seconds}.

We found one input word (\cdt{lay}) presented some very unusual behavior. Entry rates had an error rate at $89.5\%$ while the WORD mean is $20.1\%$ and no other word had an error rate higher than $30\%$. The reason is that the recogniser promotes words of higher prior probability in the language model, in this case “Larry”, “last” or “Katy” - highlight the limitations of pure gesture typing interaction.

A statistical analysis showed a significant main effect of DEVICE ($F_{1,11} = 37.77, p < 0.0001$) on error rate with mean value for TP1 and OP1 equal to respectively 6.2\% and 26.1\%, see Figure 4.a. A statistical analysis also showed a significant main effect of SIZE ($F_{2,22} = 10.99, p < 0.001$) on the error rate. Finally, an ANOVA on TP1, OP2, OP4 and OL2 did not show a significant main effect (p = 0.11) even though OPTICAL is in average higher than TABLET. In other words, among all levels of the experiment, only OP1 shows a significant higher error rate.

A statistical analysis showed a significant main effect of DEVICE ($F_{1,11} = 90.15, p < 0.0001$) on input rate with mean values for TP1 and OP1 equal to respectively 29.2 WPM and 13.7 WPM. We also looked at pair-wise comparison2 for OP-TICAL and could not find a statistical difference in the mean input rate. The input rate achieved by the participants in TP1 is in-line with what can be expected from novice users after the time of the experiments~\cite{Kristensson2004}. The averaged 54\% lower input rate in OPTICAL should be compared with other simi- lar results, as~\cite{Markussen2014} with 57\% after 10 sessions, that compare direct and indirect input modality for mid-air gesture typing.

The participants were capable of moving their hand with great variability when it comes to speed.

The Nasa TLX reported high stress on big surfaces, and high stress on the very small surface. As a results, the optimal size is in the middle.

% \begin{figure}
% \centering     %%% not \center
% \subfigure[]{\includegraphics[width=0.49\linewidth]{figures/err_DEVICE_SHAPE.pdf}}
% \subfigure[]{\includegraphics[width=0.49\linewidth]{figures/wpm_DEVICE_SHAPE.pdf}}
% \caption{Effect of levels on error rate (a) and input rate (b).}~\label{fig:effect_shape}
% \end{figure}

% =============================================================================
\section{Conclusion and Future work}
% =============================================================================

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/wpm_DEVICE_SHAPE.pdf}
    \caption{Effect of levels on input rate.}
    \label{fig:wpm_DEVICE_SHAPE}
\end{figure}
}
The user study showed that the users hit the accuracy ceiling when using a small surface. It also showed that the task behave according to Fitts law. The users seem to be limited by their ability to drive an indirect pointer rather than performing the difficulty of the task itself.

These encouraging results open the way to testing the system with patients with spinal cord injury. More so, due to recent advance in hand pose recognition, the reality of interacting via touch through vision based system should be investigated further.

@@

\newpage

\balance
\bibliographystyle{acm-sigchi}
\bibliography{collection}

>>>>

\numberofauthors{4}
% Notice how author names are alternately typesetted to appear ordered in 2-column format;
% i.e., the first 4 autors on the first column and the other 4 auhors on the second column.
% Actually, it's up to you to strictly adhere to this author notation.
\author{
  \alignauthor{
    \textbf{Antoine Loriette}\\
    \affaddr{Dept. Computing Science}\\
    \affaddr{University of Glasgow}\\
    \affaddr{Scotland}\\
    \email{antoine.loriette@gmail.com}
  }\alignauthor{
    \textbf{Sebastian Stein}\\
    \affaddr{Dept. Computing Science}\\
    \affaddr{University of Glasgow}\\
    \affaddr{Scotland}\\
    \email{author5@anotherco.com}
  }
  \vfil
  \alignauthor{
    \textbf{Roderick Murray-Smith}\\
    \affaddr{Dept. Computing Science}\\
    \affaddr{University of Glasgow}\\
    \affaddr{Scotland}\\
    \email{author2@anotherco.com}
  }\alignauthor{
    \textbf{John Williamson}\\
    \affaddr{Dept. Computing Science}\\
    \affaddr{University of Glasgow}\\
    \affaddr{Scotland}\\
    \email{jhw@dcs.gla.ac.uk}
  }
\vfil
}

@@

\begin{abstract}
The association of tabletop interaction with gesture typing presents potential for physically impaired users. In this work, we use depth cameras to create touch surfaces on regular tabletops. We describe our prototype system and report on a supervised learning approach to fingertips touch classification. We follow with a gesture typing study that compares our system with a control tablet scenario and explore the influence of input size and aspect ratio of the virtual surface on the text input performance. We show that novice users perform with the same error rate at half the input rate with our system as compared to the control condition, that an input size between A5 and A4 present the best tradeoff between performance and user preference and that user's indirect tracking ability seems to be the overall performance limiting factor.

@@

Depth cameras~\cite{Harrison2011,Xiao2016,Sridhar2017} have usually been employed to create such surfaces, while other work have tried to combine different sensor sources~\cite{Wen2016}. However, the touch classification, critical to the quality of the interaction remains problematic (sub capacitive quality?) and few systems are available for all camera types.

@@

% =============================================================================
\section{System overview}
% =============================================================================
The envisioned interaction scenario is shown in ~\autoref{fig:banner} with its implementation in \autoref{fig:system_insitu_0}. We use an Intel Realsense SR300 obliquely mounted on a tripod that overlooks the interaction surface. Finger detection and tracking is performed on a desktop computer. The system implements the 3 state button model, with its touch events sent to the Android tablet running a custom application overwriding its input event system. In addition, the tablet displays hover state as a red marker at the pointer’s position, touching is displayed through the Android debugging facility as a cross spanning the screen and continuous touch is displayed as a segment line. An audio feedback is produced on each touch down event.

@@

The touch classification problem is usually solved via flood-filling~\cite{Harrison2011,Xiao2016,Sridhar2017}. However, it is well suited to a supervised learning approach provided availability of a training dataset. One of the authors video-taped 6 minutes of interactions, half touching and half hovering with some frames out of range while using the index, thumb and pinky at close and long range. From the fingertip pointcloud, we use the histogram of z-values as features (providing some orientation and shape invariance) and a neural network classifier. We tested the generalisation with different cross validation splits and obtained an averaged 0.96 AUC for the ROC, see~\autoref{fig:roc_auc}. The ROC curves led us to choose an operating point of 0.5 for the live system, which we trained on the whole dataset for 75 epochs.

@@

% =============================================================================
\section{User study}
% =============================================================================
We ran an experiment with two research goals in mind. First we wanted to evaluate the usability of our system, for this purpose we included a condition with an interaction directly on the tablet. Second, given the nature of the task and its inherent difficulty for novice users, we investigated the influence of the physical space on performance (Accot et al.~\cite{Accot2001} demonstrated a U-shaped curve in performance against input scale). Our independent variable was thus the control surface dimensions, which is effectively changing the control-display gain ($CD_{gain}$) defined as the ratio between $V_{display}$ and $V_{control}$.

@@

We gathered 12 participants, all right handed, without requirement on gesture typing experience. A repeated measures within-subjects design was used. There were 3 conditions and 5 level combinations were evaluated, see \autoref{tab:cdt} below for details. The presentation order of the 5 combinations was randomised and presented a uniform distribution across the participants.

The task was to write 20 words per level (words taken at random from the most common english words with length between 2 and 5 letters) with maximum 7 attempt to complete 5 correct input for each word. The design of the task is similar to \cite{Quinn2016} and allows novice users to focus on the physical execution instead of the shape recollection, effectively emulating a proficient bahavior even for novice users. After each level, participants were offered to take a break before moving to the next one. Finally, participants were asked for their feedback using the NASA Task Load Index~\cite{Hart1988} (NASA-TLX) to assess the perceived workload of completed level.

@@

% =============================================================================
\section{Result}
% =============================================================================
The dependent variables are the success rate, the time taken per trial and the trace data when available. This allowed us to compute the dependent measure error rate defined as the percentage of unsucessful attempts as well as the text entry rate measured in Words Per Minute (WPM), as in [3], according to the following formula\footnote{$WPM = |T|/s \times 60/5$ where $|T|$ is the length of the transcribed string, s is time in seconds}.

We found one input word (\cdt{lay}) presented some very unusual behavior. Its error rate was at $89.5\%$ while the \cdt{WORD} mean is $20.1\%$ and no other word had an error rate higher than $30\%$. The reason is that the recogniser promotes words of higher prior probability in the language model, \textit{``Larry''}, \textit{``last''} or \textit{``Katy''}. For the all subsequent analysis of error rate, \cdt{lay} is removed from the dataset.

A statistical analysis showed a significant main effect of DEVICE ($F_{1,11} = 37.77, p < 0.0001$) on error rate with mean value for \cdt{TP1} and \cdt{OP1} equal to respectively 6.2\% and 26.1\%, see \autoref{fig:err_DEVICE_SHAPE}. A statistical analysis also showed a significant main effect of \cdt{SIZE} ($F_{2,22} = 10.99, p < 0.001$) on the error rate. Finally, an ANOVA on \cdt{TP1}, \cdt{OP2}, \cdt{OP4} and \cdt{OL2} did not show a significant main effect ($p = 0.11$) even though \cdt{OPTICAL} is on average higher than \cdt{TABLET}. In other words, among all levels of the experiment, only \cdt{OP1} shows a significant higher error rate.

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{figures/wpm_DEVICE_SHAPE.pdf}
    \caption{Effect of levels on input rate.}
    \label{fig:wpm_DEVICE_SHAPE}
\end{figure}
}

A statistical analysis showed a significant main effect of \cdt{DEVICE} ($F_{1,11} = 90.15, p < 0.0001$) on input rate, see \autoref{fig:wpm_DEVICE_SHAPE}, with mean values for \cdt{TP1} and \cdt{OP1} equal to respectively $29.2 WPM$ and $13.7 WPM$. We also looked at pairwise comparison\footnote{Post-hoc analysis were adjusted using Bonferroni correction.} for \cdt{OPTICAL} and could not find a statistical difference in the mean input rate. The input rate achieved by the participants in \cdt{TP1} is in-line with what can be expected from novice users after the time of the experiments~\cite{Kristensson2004}. The averaged 54\% lower input rate in \cdt{OPTICAL} should be compared with other similar results, as~\cite{Markussen2014} with 57\% after 10 sessions, that compare direct and indirect input modality for mid-air gesture typing.

\subsection{Effect of orientation}
The design of the experiment also includes \cdt{ORIENTATION}. This condition has so far been excluded from the SIZE analysis and can be difficult to apprehend since not only the \cdt{ORIENTATION} of the visual feedback is different but its size also. The main result is the average pointer speed in display space at $303.75 pixel/s$, higher than the PORTRAIT orientation at an average $228 pixel/s$. It is important to keep in mind that the keyboard area in LANDSCAPE is $61cm^2$, bigger than PORTRAIT at $44cm^2$, which can explain the faster pointer speed but is not responsible for a higher input rate.

\subsection{Qualitative data}
At the end of the experiment, participants were invited to provide some informal feedback, rank the different level in order of preference and fill in the NASA-TLX form, the tablet interaction was ranked best by all the participants except one. \cdt{OP4} was consistently ranked last, while \cdt{OP2} and \cdt{OL2} had equal ranking in second position. The NASA-TLX data shows that participants describe \cdt{OP4} as the most physically demanding interaction. Participants graded \cdt{OP1} and \cdt{OP4} 20\% higher than \cdt{OP2} and \cdt{OL2} on the scale of effort and frustration. Finally, \cdt{OP1} was graded as having lowest level of performance and highest mental and temporal demand.

\section{Discussion}

We proposed a machine learning approach to fingertip touch detection. This problem is traditionally addressed by hand-tuning parameters and thresholding. In contrast, we showed that using a machine learning approach can solve the problem and generalise well enough to be used in an experiment with 12 unseen participants. One improvement could be to delegate the feature extraction step (an histogram of z-values in our case) to a model capable of inferring features directly, such as a convolutional neural network. Since the data is cheap to record, we could see this approach being used for many specific cases.

We conducted a user study investigating the usability of the system and showed that for control dimension at least twice the display dimension novice users were capable the same error rate at half the input rate. We also showed that at the same $CD_{gain}$, the error rates were dramatically higher. Also, the experiment showed a constant input rate across display sizes and aspect ratios. Accot et al.~\cite{Accot2001} showed that task with high index of difficulty would exhibits a u-shaped curved with size. Since we did not observe such a behavior, this puts in question whether gesture typing for novice users is “hard enough”.

The constant input rate however shows that participants are capable of adapting their motor speed across all investigated scales. Participants also varied the display pointer speed, especially in \cdt{OL2} when presented with a bigger display surface. Because participants are capable of adapting their motor behavior and their control bahavior, another explanation for the upper bound in text input is that the indirect nature of the interaction is the limiting factor. (more research in that direction?)

From the participants feedback, it shows that \cdt{OP1} was a demanding interaction. It also showed that increasing the surface size to the extent of \cdt{OP4} is physically demanding. Recommended sizes for interaction will thus be \cdt{OP2} or \cdt{OL2} (note that the performance is also dependent on the $CD_{gain}$). Moreover, some participants reported the texture having an impact of the interaction. Levesque et al.~\cite{Levesque2011} have used friction in a dynamic manner to improve target acquisition. Given the relatively low performance of the participants, it could be interesting to explore if tactile feedback could help controlling an indirect pointer.

% % =============================================================================
% \section{Conclusion and Future work}
% % =============================================================================

% The user study showed that the users hit the accuracy ceiling when using a small surface. It also showed that the task behave according to Fitts law. The users seem to be limited by their ability to drive an indirect pointer rather than performing the difficulty of the task itself.

% These encouraging results open the way to testing the system with patients with spinal cord injury. More so, due to recent advance in hand pose recognition, the reality of interacting via touch through vision based system should be investigated further.

@@

% \newpage

% \balance
\bibliographystyle{acm-sigchi}
\bibliography{collection}
